How to write a PhD in Biological Sciences: a guide for the uninitiated

### 2024-09-11 1.04-email.Rmd
Add files via upload
# Changes to the book since April 2024 with several new chapters & core-skills as Part 1

# Communicating by email {#email}

Throughout your PhD, you will be expected to communicate with your advisor and other professionals inside and outside of academia using email. The chances are that you have already written a great deal of emails in the past. However, this does not mean that the style and content of those emails will be appropriate when communicating during your postgraduate studies.

@@

## The subject
Do include a succinct subject for your email. Do not leave this blank. Remember that your advisor, and other professionals that you communicate with, likely receives many emails in a day. They may not be able to respond to you immediately, but may want to know what the email is about before they open it.

@@

## Writing style
In general, it’s best to keep your style brief. If you feel that you need to write a long email, consider ways in which you could better communicate the same information in a meeting, or in responses to comments in a chapter draft. I would discourage you from writing long emails as reading them will tend to take up more of your advisor’s time than other forms of communication.

@@

 - try to make the reason why you are communicating clear
 - read through what you've written before you press send
 - let brevity be your watchword

### Don't use slang or SMS language
I remember an email that a colleague received (many years ago) asking for a bursary. I don't remember the exact wording, but it went something like this:

@@

### Formalities {#formal}
The level of formality that you use will largely depend on your existing relationship with your correspondent. The formal tone is set immediately with the greeting that you use.

@@

## Why exactly are you communicating?
The first two or three sentences of your email should state why you are communicating and whether or not you are expecting a reply. Emails can be just for information, keeping someone up to date on your progress or letting them know that something they asked for has been done. If this is the case, then start your email with a statement letting the recipient know the reason why you are keeping them informed, and that you don’t expect a reply.

@@

### Using Structure
I like it when people use some structure in their email. If it’s going to be a long email (which I’d advise against) then do include a short summary explanation at the start so that the reader knows what the email is about (especially if the subject line doesn’t fully explain). Using bullet points or numbered points in the email is also a great way of using structure, especially when you expect a reply on each point.

#### Replying inline
This simply means that the previous email is reproduced when you hit reply, often with some marker for what was written originally. I like inline replies as it’s simple to see that each point has been addressed. It’s also a useful way to be reminded of what was originally written.

@@

#### HTML emails or plain text
Most email software now defaults to emails written in HTML, which means that you can use **bold**, *italics*, bullet points and most other features that you would in a word processing package. Some people prefer plain text only, but HTML is becoming more standard.

#### Email chain
Most email software has an option to include the original message in the reply. This is quite useful in that it can remind everyone about what was asked. As these kinds of conversations get longer, they can become very large and unwieldy.

#### Bullet pointed or numbered lists

@@

### Use of all Capital Letters - DON’T DO THIS!!!!

@@

## Signing off {#signoff}

@@

## Who else should your email be sent to?

@@

## When to CC someone

@@

### Use of BCC

@@

## Reply or Reply All? {#replyall}

@@

### High Importance
Some email software has the ability to flag an email as important. Avoid using this unless you have something with real importance, like some decision or response needed that same day. Otherwise, try not to use this feature.

### Sending emails outside office hours
Personally, I have no problem with people sending emails whenever they want, as long as they don't expect replies to come out of office hours. I have often noted that chapter drafts appear in my inbox late on a Friday night. My suggestion to students is that this is not a good use of their time. Rather keep the draft and read it later on Sunday evening themselves after having left it alone for the weekend. You are more likely to spot errors and improve your own writing this way. It will certainly be better than it sitting untouched in my inbox all weekend.

@@

### Send Later {#later}

@@

Another scenario that I use for send later is when I have a deadline, but am awaiting feedback. I can make sure that I meet the deadline by sending the email with report or manuscript attachment or whatever it is that I have to send before the deadline. However, if at the 11th hour I get the feedback from the third party that I'm hoping to include, I can still find the email with report (it's usually in the Draft folder), make any amendments and then send it off.

### Forwarding emails

@@

## Out of Office autoreply
I find the autoreply saying that someone is away very useful. It lets you know whether or not to expect a reply, and might mean that you then write to someone else instead. If setting your own Out of Office autoreply:
 - Do include the date when you will be back (and check that it’s correct when you send it)
 - Do state whether you are on holiday or in the field 
 - If there is another means of reaching you (when you aren’t on holiday) then give this

@@

## Using email signatures and banners

@@

## Using emails as a record of due diligence {#deliverables}

@@

**If you are communicating with your advisor on a web based platform (like Slack), then do you need to send them emails as well?**

@@

## Keeping emails

@@

### Folders

@@

>>>>

# Communicating by email and messaging apps {#email}

Throughout your PhD, you will be expected to communicate with your advisor and other professionals inside and outside of academia using email. In some contexts, you may also be expected to use social media apps like WeChat and WhatsApp to communicate either directly or in groups. The chances are that you have already written a great deal of emails and instant messages in the past. However, this does not mean that the style and content of those emails and app posts will be appropriate when communicating during your postgraduate studies.

In this chapter I will start by considering some issues around correct use of email, and then move onto direct messaging apps [below](#wechat).

## Email

@@

### The subject
Do include a succinct subject for your email. Do not leave this blank. Remember that your advisor, and other professionals that you communicate with, likely receives many emails in a day. They may not be able to respond to you immediately, but may want to know what the email is about before they open it.

@@

### Writing style
In general, it’s best to keep your style brief. If you feel that you need to write a long email, consider ways in which you could better communicate the same information in a meeting, or in responses to comments in a chapter draft. I would discourage you from writing long emails as reading them will tend to take up more of your advisor’s time than other forms of communication.

@@

 - try to make the reason why you are communicating clear
 - read through what you've written before you press send
 - try not to be overly familiar or informal [@lewin-jones2014understanding]
 - be polite, as this may make a difference to you getting a positive response [@bolkan2012you]
 - let brevity be your watchword

#### Don't use slang or SMS language
I remember an email that a colleague received (many years ago) asking for a bursary. I don't remember the exact wording, but it went something like this:

@@

#### Formalities {#formal}
The level of formality that you use will largely depend on your existing relationship with your correspondent. The formal tone is set immediately with the greeting that you use.

@@

### Why exactly are you communicating?
The first two or three sentences of your email should state why you are communicating and whether or not you are expecting a reply. Emails can be just for information, keeping someone up to date on your progress or letting them know that something they asked for has been done. If this is the case, then start your email with a statement letting the recipient know the reason why you are keeping them informed, and that you don’t expect a reply.

@@

#### Using Structure
I like it when people use some structure in their email. If it’s going to be a long email (which I’d advise against) then do include a short summary explanation at the start so that the reader knows what the email is about (especially if the subject line doesn’t fully explain). Using bullet points or numbered points in the email is also a great way of using structure, especially when you expect a reply on each point.

##### Replying inline
This simply means that the previous email is reproduced when you hit reply, often with some marker for what was written originally. I like inline replies as it’s simple to see that each point has been addressed. It’s also a useful way to be reminded of what was originally written.

@@

##### HTML emails or plain text
Most email software now defaults to emails written in HTML, which means that you can use **bold**, *italics*, bullet points and most other features that you would in a word processing package. Some people prefer plain text only, but HTML is becoming more standard.

##### Email chain
Most email software has an option to include the original message in the reply. This is quite useful in that it can remind everyone about what was asked. As these kinds of conversations get longer, they can become very large and unwieldy.

##### Bullet pointed or numbered lists

@@

#### Use of all Capital Letters - DON’T DO THIS!!!!

@@

### Signing off {#signoff}

@@

### Who else should your email be sent to?

@@

### When to CC someone

@@

#### Use of BCC

@@

### Reply or Reply All? {#replyall}

@@

### Institutional or personal email addresses

As a student at your institution, you will likely be allocated an institutional email address. You probably also have your own private email addresses (e.g. Gmail, Hotmail, Yahoo, etc.). Some institutions have great problems with their mail servers, and many academics will abandon their use of their institutional email address because it causes too much trouble for them.

It is worth asking your advisor which address they would like you to use (if you know that they have several). However, I would suggest that you make use of the institutional address for all formal communication so that you have a record of [due diligence](#deliverables) should you need it.

#### High Importance
Some email software has the ability to flag an email as important. Avoid using this unless you have something with real importance, like some decision or response needed that same day. Otherwise, try not to use this feature.

#### Urgent emails {#urgent}
Sometimes it is necessary to communicate urgently. Deadlines are unforgiving and your advisor and other people are not always aware of when they are or what they mean to you. Deadlines are also very stressful, and you might be more prone to miscommunicating or making mistakes.

If you need to communicate urgently by email, then here are some tips about how to get a swift response:

1. **Urgent for whom?** I think it's worth starting by asking yourself why exactly this message is urgent, and for whom. It could be that this is only urgent for you and your deadline, or it could be that this is a shared urgency among you and the people you are corresponding with. Alternatively, the urgency might not be for you and only for the other person/people. Starting out knowing this information is important as it will help you set the tone for the message. 
2. **State that it is urgent in the subject**: Start the subject with "Urgent" and give a timeline by when you need an answer in the subject line. Most people will receive their emails on their phones and at some point they will see the subject line without reading the contents. If the subject line can inform them of what the topic is and when the urgency is for (e.g., Urgent: signature needed in the next 2 hours, the next 2 days, the next 2 weeks, etc.), this will really help them knowing exactly what is needed.
3. **Within the message**, provide the explanation again together with a clear and concise account of exactly what needs to be done. If you need a signature, then remember the attachment that needs to be signed. The more concise you can be about what is needed, the faster it is likely to be done.
4. **Follow up with an instant message or text**. If this is urgent and especially if it is outside office hours, then follow up with an instant message to check that they have received your message. See [section on texting below](#WeChat).

Before you send the email, take a few moments to read it through and check that everything is correct. If it really is urgent, then the last thing you need is to have some to and fro over the actual meaning of what is written.

Some deadlines are really important, and some correspondents are famously late with responses. In this case, consider providing a deadline one or two days earlier than the actual deadline. This only works when you have a couple of weeks to go, but it might help you build in more time and reduce the urgency. For example, you can use the [Send Later](#later) feature to send a couple of reminders building up the the date. But remember to cancel them if you get what you need before!

#### Sending emails outside office hours
Personally, I have no problem with people sending emails whenever they want, as long as they don't expect replies to come out of office hours. I have often noted that chapter drafts appear in my inbox late on a Friday night. My suggestion to students is that this is not a good use of their time. Rather keep the draft and read it later on Sunday evening themselves after having left it alone for the weekend. You are more likely to spot errors and improve your own writing this way. It will certainly be better than it sitting untouched in my inbox all weekend.

@@

#### Send Later {#later}

@@

Another scenario that I use for send later is [when I have a deadline](#urgent), but am awaiting feedback. I can make sure that I meet the deadline by sending the email with report or manuscript attachment or whatever it is that I have to send before the deadline. However, if at the 11th hour I get the feedback from the third party that I'm hoping to include, I can still find the email with report (it's usually in the Draft folder), make any amendments and then send it off.

#### Forwarding emails {#fwd}

@@

### Out of Office autoreply
I find the autoreply saying that someone is away very useful. It lets you know whether or not to expect a reply, and might mean that you then write to someone else instead. If setting your own Out of Office autoreply:
 - Do include the date when you will be back (and check that it’s correct when you send it)
 - Do state whether you are on holiday or in the field 
 - If there is another means of reaching you (when you aren’t on holiday) then give this

@@

### Using email signatures and banners

@@

### Using emails as a record of due diligence {#deliverables}

@@

**If you are communicating with your advisor on a web based platform (like Slack, WhatsAPp or WeChat), then do you need to send them emails as well?**

@@

### Keeping emails {#keepemails}

@@

### Institutional Policies

Your institution is very likely to have a policy on email use. If you are unsure and there is no summary available to you from your department, you can ask for a copy of the policy to see what is officially allowed and what is prohibited.

#### Folders

@@

## Use of WeChat or WhatsApp during your PhD studies {#wechat}

[WeChat](https://wechat.com/) and [WhatsApp](https://whatsapp.com/) are two social media direct messaging apps that currently dominate the markets in Southeast Asia and the rest of the world, respectively. In terms of their potential to communicate during your PhD, they have some similarities which I will outline first, and some differences which I will treat later (see [here](#diffs)). Other platforms, such as [Slack](https://slack.com/) might be used in particular labs. These can be highly efficient ways of communicating with groups.

Just like [emails](#keepemails), these social messaging apps can be kept and archived. This means that you can use them as a means of documenting [due diligence](#deliverables). Make sure that you have your advisor's consent to do this. They, or your institutional policies, may insist that you inform them in an email.

It is important to bear in mind that these social media apps can have an adverse effect on your mental health [@montag2018multipurpose], and even become addictive [@hou2017excessive]. If you think that you need help because you are spending too much time checking your social media apps, then please get in touch with your university's psotgraduate student support.

### Direct messaging

It may be appropriate and acceptable to use a messaging app to communicate directly with your advisor. For example, if you have an [urgent request](#urgent). Many of the points discussed [above](#email) also apply here, and some are even more relevant:

 - Although a message may be received instantly, this does not mean that your advisor (or any other academic) may have the time to respond to you immediately. You need to give people time to respond. In particular this is especially important in evenings, weekends and holiday periods. 
 - As there is an expectation that direct messaging receives a quick answer, please try to avoid sending direct messages outside of office hours - unless this has been specifically sanctioned
 - If you are new at communicating with your advisor, ask them whether they prefer to use direct messaging over email. 
 - Having someone's direct messaging details means that you can call them and speak to them directly on their mobile device. Ask permission in the chat before you do this. If you are communicating with your advisor, ask them specifically about whether they are happy with this type of communication - don't just assume it's ok.
 - Rather like email, it is possible to [forward content](#fwd) from WeChat and WhatsApp to other uses. Be sure that this is appropriate and sanctioned before doing so. Alternatively, assume that everything that is shared with you is confidential unless you are given specific permission to share. 
 - Do remember to be polite and thankful when people help you through direct messages.

### Group messaging

In general, I would advise to learn the unwritten rules of the group through observation over time. Here are some issues that I have come across with student messaging groups.

Many students love to use gifs to display their emotions. I would suggest that this is inappropriate in most instances for two reasons: first they take up a lot of space and demand that anyone trying to catch up needs to scroll through a lot more content; second they are designed to be distracting and that is not the aim of communicating in these groups.

Keep your messages on topic. Because every message that you (or anyone else) sends on this group platform will send a notification to all subscribers, it is important that you keep any messages that you send relevant to everyone in the group. Avoid sending messages that are about social events (e.g. party invites) unless this has already been sanctioned and (importantly) everyone is invited.

#### In your postgraduate classes

Joining a group for your class may be mandatory in some institutions, or at the very least the only way in which you will learn about tasks and deadlines for postgraduate classes. Some postgraduate classes are experimenting with platforms that are only mediated through groups, such as Problem-Based Learning [@luo2021wechat]. In these instances you will have to use the app.

Before you post a question on the group, make sure that someone hasn't just already asked the same question. Use the search functions and see if there's anything that you have missed.

In your classes, it would be inappropriate to add someone who isn't in the class to the group. This might be seen as being nosey or snooping on class discussions.

If you are in doubt about what can and cannot be done in a class group, consult your lecturer or ask your class rep to consult them.

#### In your lab groups

Your lab group may also have a group messaging. Generally, these are extremely efficient ways to disseminate information to a group. Be aware of the general issues with these platforms and make sure that if you want to try to do something new on your lab group, have the blessing of your advisor or senior lab member before posting.

If your group message is intended for one person in particular, then rather don't bring it to the entire group. In the same way, if someone has what you believe to be inappropriate behaviour in the lab try to take this problem to the lab PI or bring it up in a lab meeting (as appropriate) rather than posting about it in the group. Calling out bad behaviour in a chat group is sometimes counter-productive and can be upsetting for many lab members.

### Desktop versions

Both WeChat and WhatsApp have desktop versions of their platforms, and these may replace the use of email in some places. For example, in China WeChat has almost entirely replaced email in many institutions such that sending an [email](#email) may elicit no response as the inbox is simply not checked. Note that this is not the case in Europe, North America and elsewhere. Institutions may well specifically forbid the official use of social media apps for official communications, so be aware of the policies at your institution.

### Turning off notifications

It is understandable that you (and others) in a group chat will want to turn off notifications if their mobile devices are pinging too often. If you do turn off your notifications for important class or lab groups, make sure that you look through the content at least once a day (excepting holidays).

### Differences {#diffs}
There are some important differences between WeChat and WhatsApp. At the time of writing, these differences are diminishing.

#### WeChat

WeChat has become a platform that offers far more than messaging and calls [@montag2018multipurpose]. You can do a lot more on it, as it effectively works as a browser. Thus within your WeChat postgraduate class group, there may be scope to include parts of the learning onto the app. All of this makes it more likely that WeChat will be used as a part of your learning experience. Indeed, there is even evidence that using WeChat responsibly for knowledge sharing at work promotes creativity [@zhang2021excessive].

Another important difference is that WeChat is often linked to someone's bank card and so can be used for transferring cash. Although this seems unlikely to be used in your PhD situation, always take care when doing this and find out before hand whether WeChat evidence for payments can be used for accounting purposes in your institution.

#### WhatsApp

At the moment, WhatsApp does not have the extended uses of the WeChat platform.

Having someone's WhatsApp contact details means that you will have their mobile phone number (or at least the one used for their account which may not be the SIM in their phone). Use this information with respect and don't share it with third parties unless you have consent.

## Other Social Media platforms

During your PhD studies you may make use of other social media platforms to communicate such as [Atlmetrics](https://howtopublishscience.org/altmetrics.html) [e.g., see @measey2022how]. As ever, use social media with caution and professionally. Respect the rights of you advisor and lab mates to have their own private social media, just as you may well have your own.

### 2024-09-11 1.05-useR.Rmd
Add files via upload
# Changes to the book since April 2024 with several new chapters & core-skills as Part 1

# Reconciling yourself to doing things you've been avoiding {#useR}

Some 30 odd years ago, as an undergraduate, I was told that I would have to use computers in one of my 2nd year courses. The thought filled me with dread and I remember putting off going into the computer room for as long as I could. It was a steep learning curve, and I struggled. As my only previous experience had been with a typewriter, each time the cursor got to the end of the screen I hit the 'carriage-return' button. And each machine had an odd white device attached called a mouse, and I had no idea what that did. All this seems quite preposterous now, but imagine embarking on a PhD in biological sciences without reconciling yourself to using a PC. This is practically where we are now with R (@rcoreteam2021language; Figure \@ref(fig:Rlogo)), and a wealth of other emerging and alternative programming platforms.

(ref:Rlogo) **The ubiquitous R logo.** R (and the invaluable GUI RStudio) has become the go to platform for many statistics and figures in biological sciences

```{r Rlogo, out.width = '33%', echo=FALSE, fig.align='centre', fig.cap="(ref:Rlogo)"}
 knitr::include_graphics("figures/Rlogo.png")
```

>>>>

# Upskilling {#useR}

Some 30 odd years ago, as an undergraduate, I was told that I would have to use computers in one of my 2nd year courses. The thought filled me with dread and I remember putting off going into the computer room for as long as I could. It was a steep learning curve, and I struggled. As my only previous experience had been with a typewriter, each time the cursor got to the end of the screen I hit the 'carriage-return' button. And each machine had an odd white device attached called a mouse, and I had no idea what that did. All this seems quite preposterous now, but imagine embarking on a PhD in biological sciences without reconciling yourself to using a computer.

It is obvious now that in order to conduct your work you are going to need skills in lots of tools. Some of these tools will be niche, but others will be universal and extend beyond your thesis into your world after graduating. The message of this chapter is that you may well find that there are skills that you need to learn during your PhD studies that you feel some initial aversion to. You may have a physical aversion to, for example, using a pipette in a lab or dissecting a euthanised animal. Or this could be some kind of skill such as programming in R or Python.

Reconciling yourself to learning new skills, even when you have an aversion, will be something that you come across time and again in life, and not just while you are doing a PhD. Because doing a PhD is a time-limited goal orientated process, you will need to quickly overcome your aversion and ulimtately this will help you in future life.

## When to invest your time in learning a new skill {#upskilling}

 Early adopters can either end up putting in a lot of effort and then finding that their effort is misplaced, or finding that they are ahead of the curve and that everyone else needs to catch up.

Knowing whether or not to invest your time in learning a new skill is difficult. I suggest that you try to find out more about exactly how much time you might need to invest, and exactly how wide or narrow this new skill might be of use in the future.

If your advisor tells you that you must learn this particular tool or skill, then you need to reconcile yourself to doing it. Remember that learning a new skill gives you more than that particular talent. Learning skills helps your brain become more flexible and adaptable to learning in the future [@chang2014reorganization].

Whatever aversion you may have to learning this new skill, you will need to approach it with professionalism, and put aside any prejudice that you may have. Use your [time management](##Time-management) skills to carve a small piece out of every day to learning your new skill. This starts with being aware when in the day you learn best, and being aware of all the other commitments you have together with how this one will rank.

### Example: Learning to write code in R
In the early 2000s, R (@rcoreteam2024language) was relatively unknown, but a few pioneering individuals had spotted the advantage to using it in their own work (especially to help with repeatability). For those early adopters I am sure that their investment in learning R has paid off.

### 2024-09-11 1.06-healthy.Rmd
Add files via upload
# Changes to the book since April 2024 with several new chapters & core-skills as Part 1

Problems arise when we become overwhelmed by stress and are unable to fully respond. When this occurs, productivity can drop off and survival responses can be triggered as if responding to an actual physical attack. These responses include fight, flight or freeze responses. Anxiety and panic can be triggered. In this state, additional demands on your time may also push your life off balance, so that you start to neglect your personal wellbeing which can negatively impact on relationships, exercise regime, or even nutrition and personal hygiene. Some people can find that the additional stress can cause physical symptoms that may even need medical treatment. Your sense of competence and mastery can be negatively impacted such that you may even suffer from feelings of inadequacy or imposter syndrome ([see part 2](#fear)).

Although there are not many studies on mental health for PhD students, those that exist (as well as surveys: [Nature 2019](https://www.nature.com/articles/d41586-019-03459-7)) all suggest that there is a significant toll, which is proportionately higher than for others in society [@levecque2017work]. Whatever your prior experience of stress in a working environment, academia is known to be particularly stressful, and as a PhD student, you are likely to absorb a significant amount of this stress into your own life [@stubb2011balancing].

@@

Right now, I suggest you complete the GHQ-12 (Table \@ref(tab:GHQ)) and record your answers as a baseline. Keep the scores somewhere safe. During the course of your PhD, if you feel that your scores may have changed, take the test again and compare them with your baseline scores. Although there are no hard rules, if three or more of your scores have moved by two or more points it could be worth discussing with your support network to help you decide whether or not to seek professional help.

@@

Academia is recognised as a particularly stressful environment; you will likely take on some of this environmental stress in addition to any stress associated with your studies. Additional stressors come from home and family situations. Your best means of coping will be to try and develop a support network and to understand where and with whom you can discuss any difficulties as they arise. Knowing who this is and how and when to approach them will put you in a stronger position if you need them in future.

@@

The positive relationship between the amount of physical activity and higher mental wellbeing is well established [e.g. @grasdalsmoen2020physical; @gerber2014increased], but the kind of exercise required to achieve this improved result is varied. For example, a study of college students showed that an hour of Thai Chi twice a week for 3 months was enough to have significantly improved physical and mental health scores [@wang2004effects]. Increased levels of physical activity were also found to be associated with improved sleep quality in another study [@ghrouz2019physical; @gerber2014increased]. There are plenty of studies out there that suggest there are multiple benefits from physical activity.

@@

There are other ways to achieve these benefits, including mindfulness or meditation.

@@

I used to think that the time I spent exercising was unproductive with respect to work. Now I have begun to realise that this represents some of my best thinking time. I use the time I spend trail running to turn ideas over in my head and especially to think through the logic in arguments, and potential flaws in experimental design. This isn’t to say that I don’t need to put time in at the computer writing it all down, or talking through ideas with colleagues. But I find the meditative time I spend exercising to be especially productive. Don't take my word for it, see this great new book by Caroline Williams [-@williams2021move].

@@

@@

Once you have a healthy body, you’ll find that there are all sorts of things that go along with this. The desire to fill it with healthy food resulted, and this later tied in with my interest in the gut microbiome. I now have many fermented foods in my kitchen (much to the horror of my kids), and the improvement of my digestive system. I could go on, but I’ll spare you. Instead, I just want to emphasise that I consider that not only will your quality of life improve, but so will your work if you include regular exercise as part of your PhD regimen.

>>>>

Problems arise when we become overwhelmed by stress and are unable to fully respond. When this occurs, productivity can drop off and survival responses can be triggered as if responding to an actual physical attack. These responses include fight, flight or freeze responses. Anxiety and panic can be triggered. In this state, additional demands on your time may also push your life off balance, so that you start to neglect your personal well-being which can negatively impact on relationships, exercise regime, or even nutrition and personal hygiene. Some people can find that the additional stress can cause physical symptoms that may even need medical treatment [see @zhong2019depressive]. Your sense of competence and mastery can be negatively impacted such that you may even suffer from feelings of inadequacy or imposter syndrome ([see part 2](#fear)). These types of pressure tend to be low at the start, and increase during your time as a postgraduate student [@cheng2020suicide].

Although there are not many studies on mental health for PhD students, those that exist (as well as surveys: [Nature 2019](https://www.nature.com/articles/d41586-019-03459-7)) all suggest that there is a significant toll, which is proportionately higher than for others in society [@levecque2017work]. Whatever your prior experience of stress in a working environment, academia is known to be particularly stressful, and as a PhD student, you are likely to absorb a significant amount of this stress into your own life [@stubb2011balancing; @cheng2020suicide].

@@

Right now, I suggest you complete the GHQ-12 (Table \@ref(tab:GHQ)) and record your answers as a baseline. Keep the scores somewhere safe. During the course of your PhD, if you feel that your scores may have changed, take the test again and compare them with your baseline scores. Although there are no hard rules, if three or more of your scores have moved by two or more points it could be worth discussing with your [support network](#support) to help you decide whether or not to seek professional help.

@@

Academia is recognised as a particularly stressful environment; you will likely take on some of this environmental stress in addition to any stress associated with your studies. Additional stressors come from home and family situations. Your best means of coping will be to try and develop a [support network](#support) and to understand where and with whom you can discuss any difficulties as they arise. Knowing who this is and how and when to approach them will put you in a stronger position if you need them in future.

## Support network {#support}

The people in your life make up your support network. Each person will have a different role, and not everyone in your support network will be able to support your mental wellbeing. For example, some of your colleagues at work might be in your support network for using equipment or doing statistics, but you may find that these sample people do not necessarily understand you as a person. People in your support network for your mental wellbeing are likely to consist of your close friends, family and those colleagues at work that you find a special connection with (i.e. new friends).

You should identify your support network, and even place a tag next to their name in your contacts. Once you know who they are, you should also let them know that they are important to you, and that you may in future come to them to discuss issues around your mental wellbeing during your PhD studies in the future. It is important that they know that they have this role for you, and more often than not you can offer to fulfil this role for them in a reciprocal manor.

During the course of your PhD, you should make sure that you take the time to check-in with your support network and let them know how things are going. This means sharing the good times as well as the bad times with them. If the only contact you make with them is when things get really bad, you are likely to find that they don't have a sufficient connection with you to help get you back on track. When you make contact with people in your support network, don't forget that they also require support from you. Inquire about how they are doing and be prepared to spend time sharing your respective highs and lows. Each time you invest time in your support network it will grow that relationship and deepen the level of support that you can hope to receive when you need it.

@@

The positive relationship between the amount of physical activity and higher mental well-being is well established [e.g. @grasdalsmoen2020physical; @gerber2014increased], but the kind of exercise required to achieve this improved result is varied. For example, a study of college students showed that an hour of Thai Chi twice a week for 3 months was enough to have significantly improved physical and mental health scores [@wang2004effects]. Increased levels of physical activity were also found to be associated with improved sleep quality in another study [@ghrouz2019physical; @gerber2014increased]. There are plenty of studies out there that suggest there are multiple benefits from physical activity.

@@

It is important to note that exercise is not about creating some perfect body image that you might see on social media. A useful tool to help think about the importance of exercise for your body came from from Stutz and Michels' [-@stutz2022tools] Life Force model in which they proposed that your life force is equivalent to a pyramid. The bottom of the pyramid represents your relationship with your body which is equivalent to 85% of the life force pyramid area (Figure \@ref(fig:life-force)). Working on this component consists of sleep, diet and exercise. The second part of this pyramid has your relationships with other people, both personal and in your work context. Lastly, your mental health (your relationship with yourself) sits atop these other components and is supported by them. Working on the lower components of the pyramid will support your mental health, but you can also work directly on your mental health through, for example, mindfulness and meditation. Stutz goes on to explain that writing is a way to enhance your relationship with yourself. This might explain why some people find writing their thesis a very positive experience, but also gives us another way in which we can think about writing and ourselves.

(ref:life-force) **A model to understand the importance of your body in connection to your mental health.** In this model from Stutz and Michels [-@stutz2022tools] you see the body taking up 85% of the area of the pyramid, on which rests your relationships with other people and finally your mental health on top. Maintaining your mental health therefore rests on your body receiving enough sleep, a good diet and exercise.

```{r life-force, echo=FALSE, out.width='50%', fig.cap="(ref:life-force)"}
 knitr::include_graphics("figures/life-force.png")
```

@@

I used to think that the time I spent exercising was unproductive with respect to work. Now I have begun to realise that this represents some of my best thinking time. I use the time I spend trail running to turn ideas over in my head and especially to [think](#thinking) through the logic in arguments, and potential flaws in experimental design. This isn’t to say that I don’t need to put time in at the computer writing it all down, or talking through ideas with colleagues. But I find the meditative time I spend exercising to be especially productive. Don't take my word for it, see this great book by Caroline Williams [-@williams2021move]. In this book it is emphasised that exercise might not mean sport, but simply taking a walk or another aspect of self propelled movement. Actively seek out the activities that work best for you in order to [think](#thinking) or [reflect](#reflection). Once you know what they are, [plan](#time) to use them to optimise your thinking time.

@@

Non-working time (e.g., holiday time) is also important in order to keep you productive during your PhD. You should not be working every day from beginning to end. Try to save at least one day every week work-free so that you can rest and spend time with friends and/or relatives. During busy times you might have to sacrifice this day, especially when you are working in the field or attending a conference. But you should make up this shortfall when times are less busy. You also need a real holiday when you take a break from your PhD work altogether at least twice every year.

@@

Once you have a healthy body, I found that there are all sorts of things that go along with this. The desire to fill it with healthy food resulted, and this later tied in with my interest in the gut microbiome. I now have many fermented foods in my kitchen, and the consequent improvement of my digestive system. I could go on, but I’ll spare you. Instead, I just want to emphasise that I consider that not only will your quality of life improve, but so will your work if you include regular exercise as part of your PhD regimen.
